---
title: "Forecasted Average Treatment Effects in the Absence of a Control Group"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasted Average Treatment Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette introduces the R package `fatEstimator`, which implements the method developed by Botosaru, Giacomini, and Weidner (2023) to estimate policy effects in the absence of a control group. The estimator relies on forecasting counterfactual outcomes based on short pre-treatment time series data.

# Theoretical Background

In standard treatment effect estimation, control groups are essential to infer counterfactual outcomes. However, many important policies (e.g., universal participation policies, fiscal shocks, large-scale mandates) lack such a group. Botosaru et al. (2023) propose the Forecasted Average Treatment Effect (FAT) estimator, which addresses this by forecasting the untreated potential outcome using the individual’s own pre-treatment history.

## Assumptions and Key Features

- **Forecast Unbiasedness**: Counterfactual forecasts must be unbiased on average.
- **Weak Dependence**: Forecast errors must satisfy a central limit theorem.
- **Basis Function Forecasting**: Forecasts are based on regressions using simple basis functions (e.g., polynomials).

## Advantages

- No control group needed.
- Allows for heterogeneous treatment effects.
- Robust to model misspecification.
- Compatible with short time series.

# Package Usage

```{r library}
#library(fatEstimation)
```

# Simulated Example

This example demonstrates how to use the package on simulated data:

```{r example}

```

# Interpretation

The function `fat_estimate()` uses a simple polynomial regression on pre-treatment data to forecast the counterfactual outcome. The ATT is then estimated as the average difference between the observed and forecasted outcomes in the post-treatment period.

# References

Botosaru, I., Giacomini, R., & Weidner, M. (2023). *Forecasted Treatment Effects*. Working Paper. Available from authors.
