---
title: "Forecasted Average Treatment Effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasted Average Treatment Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


The `fatEstimator` package implements the *Forecasted Average Treatment Effect Approach* from Botosaru, Giacomini, and Weidner (2024). This method is designed for settings without a reliable control group, such as universal policy interventions, endogenous treatment adoption, or when traditional difference-in-differences assumptions are not credible. Additionally, the approach can be a useful robustness check for settings with a control group, especially when its validity is questionable or the assumptions of the existing methods are too strong to be credible.

Rather than comparing treated units to untreated ones, the FAT method builds unit-specific forecasts of the counterfactual outcome (i.e., what would have happened without treatment), based only on pre-treatment data. The treatment effect is then estimated as the difference between the observed outcome and this forecast. By focusing on unit-specific forecasts, the method allows for heterogeneous treatment effects, unbalanced panels, and staggered (or individual-specific) treatment adoption.

This vignette walks you through:

- The baseline estimator based on pre-treatment trend forecasting,
- The theoretical motivation and required assumptions,
- And model extensions (covariates, heterogeneity, control groups).


## 1. Estimation Strategy: Baseline Case

Assume that all units are treated at a common time $\tau$. The treatment effect at $h$ periods after treatment is defined as:

$$
\widehat{FAT_h} := \frac{1}{n} \sum_{i \in n} [y_{i \tau + h} - \widehat{y}_{i \tau + h}(0)] \tag{1}
$$

- $Y_{i \tau + h}$ is the observed outcome for unit $i$ in period $\tau + h$,
- $\widehat{Y}_{i \tau + h}(0)$ is the forecasted counterfactual - what would $Y$ have been in the absence of treatment.

This estimator does not require a control group. Instead, the counterfactual $\widehat{Y}_{i,\tau+h}(0)$ is predicted using only pre-treatment observations for unit $i$. 


### 1.1 Forecasting the Counterfactual

The counterfactual $\widehat{y}_{i, \tau + h}(0)$ is constructed using a **basis function regression** fit to the $R_i$ periods prior to treatment with $\tau$ denoting the time of the treatment. The method requires specifying:

- A set of linearly independent basis functions $\{ b_k(t) \}_{k=0}^{q_i}$, here polynomial basis functions $b_k(t) = t^k$ are used by default.
- The number of basis functions $q_i$
- The estimation window $T_i = \{ \tau - R_i + 1, \dots, \tau \}$


$$
\widehat{y}_{i \tau + h}^{(q_i, R_i)}(0) = \sum_{k=0}^{q_i} \widehat{c}_{ik}^{(q_i, R_i)} (\tau + h)^k \tag{2}
$$

The coefficients $\widehat{c}_i = (\widehat{c}_{i0}, ..., \widehat{c}_{iq_i})$ are estimated by minimizing the squared error over the pre-treatment window:

$$
\widehat{c}^{(q_i, R_i)}_i = \arg\min_{c \in \mathbb{R}^{q_i+1}} \sum_{t \in T_i} \left(y_{it} - \sum_{k=0}^{q_i} c_k t^k\right)^2 \tag{3}
$$

### 1.2 Theoretical Justification

Under suitable conditions on the data generating process (DGP), the forecast $\widehat{y}_{i, \tau + h}(0)$ is **unbiased on average**, , i.e.: $\mathbb{E}[\widehat{Y}_{i, \tau + h}(0)] = \mathbb{E}[Y_{i, \tau + h}(0)]$, so that the FAT estimator consistently estimates the ATT. 

Thus, the FAT estimator consistently estimates ATT. See Theorem 1–2 in Botosaru et al. (2024).

In particular:

- If the DGP includes **stationary or stochastic trends**, any weighted average of pre-treatment outcomes (with weights summing to 1) is unbiased (Theorem 1).
- If the DGP includes **deterministic trends**, basis function regressions remain unbiased provided the true trend is contained in the span of $\{ b_k(t) \}$ and $q_i \geq q_{0i}$ (Theorem 2).

Thus, polynomial regressions of modest order (e.g., $q_i = 0, 1, 2, 3$) offer a flexible and theoretically valid basis for forecasting the counterfactuals.

#### Assumptions and Key Features

- **Forecast Unbiasedness**: Counterfactual forecasts must be unbiased on average.
- **Weak Dependence**: Forecast errors must satisfy a central limit theorem.
- **Short Panel**: Identification and inference do not require long time series. 
- **Unbiased Forecasts**: A correctly specified model is not necessary to obtain unbiased forecasts of the counterfactuals. 

#### Practical Advice
- Choose $q_i$ (polynomial degree) based on data visualization. Use 0 for constant trend, 1 for linear, etc.
- Choose $R_i$ (length of pre-treatment window) by balancing the desire for small standard errors (large $R_i$) with concerns about structural instability pre-treatment (small $R_i$).
  - The smallest $R_i$ that allows estimating the polynomial degree $q_i$ is $R_i = q_i + 1$, which is also set as the package default. 
- Placebo checks: Apply the FAT estimator to a pretend treatment **before** the actual implementation. If estimates are statistically near zero, the unbiasedness assumption cannot be rejected, and there is indication that FAT estimates the causal effect of the treatment.
- Be cautious: this baseline case does not account for:
  - Common unforecastable shocks,
  - covariates (e.g., time-varying factors),
  - or staggered treatment with anticipation. 


## 2. Model-based FAT

The FAT estimator can be extended to handle more complex scenarios. This requires specifying a model for the counterfactuals and imposing additional assumptions.

## 2.1 Homogeneous Coefficients

To incorporate additional covariates $x_{it}$ (e.g., policy drivers, controls), we forecast:

$$
\widehat{y}_{h}(\beta, y_i, x_i) = x_{i \tau + h}' \widehat{\beta} + \sum_{k=0}^{q_i} \widehat{c}_{ik} (\tau + h)^k \tag{5}
$$
$$
\widehat{c_i}^{(q_i, R_i)} = \arg\min_{c \in \mathbb{R}^{q_i + 1}} \sum_{t \in T_i} \left( y_{it} - x_{it}'\widehat{\beta} - \sum_{k=0}^{q_i} t^k c_k \right)^2 \tag{6.1}
$$


The coefficients $\beta$ are common across units, while $c_{ik}$ remain unit-specific. The $\widehat{c}_i$ are obtained by regressing $y_{it} - x_{it}'\widehat{\beta}$ on $\{t^k\}$ over $t \in T_i$. The $\widehat{\beta}$ can be estimated using pooled OLS or IV (see below).

This is implemented in the package using `beta_estimator = "ols"` or `"iv"`.

#### Use Case
Use this when some covariates (e.g. exogenous controls, lagged outcomes, or instruments) explain common dynamics.

#### Requirements
- $x_{it}$ must be observed in all periods.
- Instruments must be supplied if `beta_estimator = "iv"`.


## 2.2 Heterogeneous Coefficients (Unitwise Estimation)

When units are highly heterogeneous, it may be inappropriate to pool data. The package allows estimating the polynomial trend *separately for each unit* without assuming common slopes.

The model is:

$$
y_{i \tau + 1}^{(q_i, R_i)}(0) := \sum_{k=0}^{q_i} c_{ik}^{(q_i, R_i)} (\tau + 1)^k + \widehat{\beta}^{(i)} x_{i \tau + 1} \tag{6}
$$


$$
(\widehat{c}^{(q_i, R_i)}, \hat{\beta}^{(i)} := \arg\min_{c \in \mathbb{R}^{q_i + 1}} \sum_{t \in T_i} \left( y_{it} - \sum_{k=0}^{q_i} c_k t^k - \widehat{\beta} x_{it} \right)^2 \tag{6.1}
$$

where $\widehat{\beta}_i$ is the unit-specific coefficient vector estimated from the pre-treatment data.

This is the default behavior if `beta_estimator = "unitwise"`.

#### Use Case
Use when units have idiosyncratic dynamics, especially when time series are long enough to estimate per-unit trends.


#### Additional Assumptions

The model-based approach requires stronger assumptions:

- correct specification of the model
- availability of a consistent estimator for the homogenous coefficients,
- symmetry of the error term for heterogeneous coefficients and covariates that are lagged outcomes. 

Under these additional assumptions, the model-based FAT is biased, but remains consistent and asymptotically normal.

## 3 Difference-in-FAT (DFAT)

Some common shocks may affect all units post-treatment. If untreated units are available, we can estimate a differential FAT:

$$
DFAT_h := FAT_h^{treated} - FAT_h^{control} \tag{7}
$$

The forecast models are fitted separately for both treated and untreated units, and their average deviations from observed outcomes are differenced.

This is implemented by supplying a `treated` column and setting `control_group_value = FALSE`.

#### Use Case
Use DFAT when:

- External shocks may bias post-treatment outcomes,
- A control group is available but not suitable for pre-treatment matching.


## 4. Package Usage

```{r library}
# library(fatEstimator)
```

### Application Steps

1. Prepare a long-format dataset with:
   - unit, time, outcome, treatment year, and optionally covariates.
2. Call `estimate_fat()` with appropriate arguments.
3. Interpret:
   - `results$results`: FAT estimates and standard errors for (degree, horizon),
   - `results$predictions`: unit-level observed and forecasted values, useful for plotting and diagnostics,
   - `plot_fat()`: visual summary of FAT,
   - `plot_fat_dfat_trajectories()`: visualizes the pre- and post-treatment of treated (and if applicable) control units over time.

## 5. Simulated Example (to be inserted)


```{r, eval=FALSE}
# Simulated code here
```


### Key user input

To implement the baseline FAT estimator, the researcher must define:

- *unit_var*: identifier for each observational unit (e.g., municipality or region).
- *time_var*: the time dimension (e.g., "year").
- *outcome_var*: the dependent variable to forecast.
- *treat_time_var*: treatment adoption year per unit.
- *degrees*: a vector of polynomial degrees (e.g., 0:2) to fit as pre-trends.
- *horizons*: a vector of forecast horizons after treatment (e.g., 1:3).
- *beta_estimator* = "none", "ols", "iv", "unitwise": 
  - "none" for no covariates,
  - "ols" for pooled polynomial fit,
  - "iv" for instrumental variable estimation,
  - "unitwise" for unit-specific trends.


# References

Botosaru, I., Giacomini, R., & Weidner, M. (2024). Forecasted Average Treatment Effects in the Absence of a Control Group. Journal / arXiv, 2024.
(Ensure correct citation details once journal info is finalized.)
Paper version used: 2024 working paper. 
